Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-04 paper/main.tex
more work torwards popl deadline

$|[x, n]| + |[y, m]|$
\\[.7\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.7\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y = y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
   \end{lstlisting}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%

@@

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume for now that a
program state $\state$ simply maps variable to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations then the cost of
the following example can be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
Also note that there are no restrictions on the signs of the input
variables in the examples.

More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other tool for C is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound $x + y + 10$, Rank~\cite{AliasDFG10} computes the
bound $y-x-7$, and LOOPUS~\cite{SinnZV14} computes the bound
$y-(x+9)$.  Only PUBS~\cite{AlbertAGPZ12} computes the tight bound
$0.1(y-x)$ if we translate the program into a term-rewriting system by
hand.

It might be surprising that we only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous work
on bound analysis~\cite{GulwaniMC09} and demonstrate the our method
can handle \emph{tricky iteration patterns}.  The SPEED
tool~\cite{GulwaniMC09} derives the same bounds as our analysis but
requires inference of \emph{disjunctive invariants}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, with the amortized analysis in mind,
using the potential transfer reasoning, it is almost trivial to prove
the tight bound $|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequential loops or function calls} through
size changes of variables.  We accurately track the size change of $y$
in the first loop by transferring the potential $0.1$ form $|[y,z]| to
|[0,y]|$.  Furthermore, \emph{t08a} shows that we do not handle the
constants $1$ or $0$ in any special way.  In all examples you could
replace $0$ and $1$ with other constants like we did in the second
loop and still derive a tight bound.  The only information, that the
analyzer needs is $y \geq c$ before assigning $y = y - c$.

Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  The outer loop we increment the variable $n$
till $n = 0$.  In each of the $|[n,0]|$ iterations, we increment the
variable $y$ by $1000$.  We then non-deterministically execute an
inner loop that decrements $y$ by $100$ until $y<100$.  The analysis
can derive that only the first execution of the inner loop depends on
the initial value of $y$.  We again derive tight constant factors.
%

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system is the only available system that can compute a tight
bound on this example.  The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

We only show a small selections of the programs that we can handle
automatically here.  In \pref{app:cat} is a list of more than $30$
classes of challenging programs that we can automatically analyze.
\pref{sec:exper} contains a more detailed comparison with other tools
for automatic bound derivation.

\begin{figure}[t]
  \centering
    \begin{minipage}[b]{\linewidth}
    \begin{center}
   \begin{lstlisting}
void count_down (int x,int y) {
  if (x>y) { tick(1); count_up(x-1,y); }
}
void count_up (int x, int y) {
  if (y+1<x) { tick(1); count_down(x,y+2); }
}
   \end{lstlisting}

$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
\\[.7\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
  \caption{Two mutually-recursive functions with the computed tick bounds.}
  \label{fig:ex_rec}
\end{figure}

@@

@@

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} that maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

For simplicity, we assume that local variables and global variables
are always disjoint.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$.  Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
For a program state $\state$, we similarly write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.

@@

\end{mathpar}
\caption{Operational semantics of statements}
\label{fig:opsem}
\end{figure}

\pref{fig:opsem} contains the reduction rules of the semantics.  The
rules define a rewrite system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost \in \Q$ is the non-negative number of resources available for
further execution.  All rules that can decrease $\cost$ have the
implicit side condition that the resource quantity available
\emph{before} the step is non-negative.  This means that we allow
$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
transition relation $\smallstep{M}^*$ to indicate that the execution
ran out of reassesses.  However, every execution that reaches such a
state is stuck.

A continuation $\cont$ represents the context of the execution of a
given statement.  A continuation can be the empty continuation
$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
\cont$, a loop continuation $\Kloop S \cont$, or the continuation
$\Kcall r \theta \cont$ of a function call.
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$
%
During the execution we assume a fixed function context $\fenv$ that
maps function names to a list of variables and the statement that
defines the function body.  It is used in the rule \textsc{S:Call}.

The intuitive meaning of an evaluation judgement $(\state, S, \cont,
\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
following.  If the statement $S$ is executed in program state
$\state$, with continuation $\cont$, and with $\cost$ resources
available then---after a finite number of steps---the evaluation will
reach the new machine state $(\state', S', \cont',\cost')$ and there
are $\cost'$ resources available.  If $\cost' \ge 0$ then the
execution did not run out of resources and the resource consumption up
to this point is $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.  If however
$\cost' < 0$ then the execution ran out of resources and is stuck.
The cost of the execution is then $\cost \ge 0$.

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
\itemskip
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
\itemskipIn
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\itemskip
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}

>>>>

$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.4\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y = y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
   \end{lstlisting}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.4\baselineskip] 
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.4\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Derivations of bounds on the number of ticks for challenging
  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
  \emph{t08a} contains \emph{sequential loops} so that the iterations
  of the second loop depend on the first, and \emph{t27} contains
  interacting \emph{nested loops}. In the potential functions, we only
  mention the non-zero terms and in the logical context $\Gamma$ we
  only mention assertions that we actually use in the reasoning. In
  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
  0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%

@@

To enable automation, we fix the shape of the potential functions
$\Phi$ so that it becomes possible to use \emph{linear programming}
(LP) to compute a derivation in the logic.  If we assume for now that
a program state $\state$ simply maps variable to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} \iffull{and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$. Note
that there are no restrictions on the signs of the input variables in
the examples.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variables for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear inequalities on
these variables that enforce that the potential annotations are sound.
For instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous
work~\cite{GulwaniMC09} and demonstrate the our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires inference of
\emph{disjunctive invariants}.  In the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, using potential transfer reasoning of
amortized analysis, it is straightforward to prove the tight bound
$|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequential loops} through size changes of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  We then non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis is the only available system that computes a tight
bound on this example.  \iffull{The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.}{}
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

\iffull{We only show a small selections of the programs that we can handle
automatically here.}{}  In \pref{app:cat} is a list of more than $30$
classes of challenging programs that we can automatically analyze.
\pref{sec:exper} contains a more detailed comparison with other tools
for \iffull{automatic}{} bound derivation.

\begin{figure}[t]
  \centering
    \begin{minipage}[b]{\linewidth}
    \begin{center}
   \begin{lstlisting}
void count_down (int x,int y) {
  if (x>y) { tick(1); count_up(x-1,y); } }

void count_up (int x, int y) {
  if (y+1<x) { tick(1); count_down(x,y+2); } }

   \end{lstlisting}

$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
\\[.4\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
\vspace{-1.5ex}

  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
  \label{fig:ex_rec}
\end{figure}

@@

\end{mathpar}
\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
\label{fig:opsem}
\end{figure}

@@

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

For simplicity, we assume that local variables and global variables
are always disjoint.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$. \iffull{Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$}{}
\iffull{For a program state $\state$, we}{We} similarly write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.

@@

\pref{fig:opsem} contains \ifshort{selected}{} the reduction rules of the semantics.  The
rules define a rewrite system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost \in \Q$ is the non-negative number of resources available for
further execution.  All rules that can decrease $\cost$ have the
implicit side condition that the resource quantity available
\emph{before} the step is non-negative.  This means that we allow
$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
transition relation $\smallstep{M}^*$ to indicate that the execution
ran out of reassesses.  However, every execution that reaches such a
state is stuck.

\iffull{
A continuation $\cont$ represents the context of the execution of a
given statement.  A continuation can be the empty continuation
$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
\cont$, a loop continuation $\Kloop S \cont$, or the continuation
$\Kcall r \theta \cont$ of a function call.
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$
%
During the execution we assume a fixed function context $\fenv$ that
maps function names to a list of variables and the statement that
defines the function body.  It is used in the rule \textsc{S:Call}.
}{}

The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
following.  If the statement $S$ is executed in program state
$\state$, with continuation $\cont$, and with $\cost$ resources
available then---after a finite number of steps---the evaluation will
reach the new machine state $(\state', S', \cont',\cost')$ and there
are $\cost'$ resources available.  If $\cost' \ge 0$ then the
execution did not run out of resources and the resource consumption up
to this point is $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.  If however
$\cost' < 0$ then the execution ran out of resources and is stuck.
The cost of the execution is then $\cost \ge 0$.

\iffull{
\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
\itemskip
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
\itemskipIn
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\itemskip
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}
}{}
